---
title: Data
description: We describe the sources of our data and the cleaning process.
toc: true
draft: false
---

### Data Background
This data comes from the Institute for Health Metrics and Evaluation and is supported by the Global Health Data Exchange.  It presents maternal mortality ratios for women in the United States from the ages 10-54 for the years 1999-2019. 

Source: https://doi.org/10.6069/35ZC-FM02

There was no clear information on the original way of collecting the data, but from the organization, it can be inferred that the data could be mostly from medical records across the country. The original guidebook directs to: https://ghdx.healthdata.org/record/ihme-data/united-states-maternal-mortality-by-state-race-ethnicity-1999-2019#:~:text=Data%20Release%20Information%20Sheet

The dataset includes MMR (maternal mortality ratio) estimates for Hispanic and any race; non-Hispanic American Indian and Alaska Native; non-Hispanic Asian, Native Hawaiian, or Other Pacific Islander; non-Hispanic Black; and non-Hispanic White females ages 10-54 for each year from 1999 through 2019.

Although the data is quite accessible,  however, pivoting the data or making subtables would likely aid in our analysis given its current organization. Some questions we can look at with this data are: “How does maternal mortality change over time? How does maternal mortality vary across US states? How does maternal mortality vary across ethnic groups?” At a deeper level, we could investigate how maternal mortality varies across both state and race groups, and compare with additional data about healthcare for different racial groups to add on a new data set. A possible challenge comes in interpreting the value for the maternal mortality ratio.

The data is often used to help in understanding the disparities in maternal health outcomes among different racial and ethnic groups. There are some publications related to this data already and they both highlight the points out the fact that maternal mortality rates are highest for the black population and the large increase trend across all groups especially in American India. One constantly asked question is how the huge increase in maternal mortality rate has happened with the development of medical technology over the past 20 years.

### Data Loading and Cleaning

[Data Loading and Cleaning Script](scripts/load_and_clean_data.R)

The file we chose initially has columns measure_id, measure_name, location_id, location_name, race_group, sex_id, age_group_id, age_group_name, year_id, metric_id, val, lower, and upper. It has 5901 observations. This odd number indicates that there may be some missing observations for certain location/race/year combinations. To load the data, we made an account and downloaded the .csv file called IHME_USA_MMR_STATE_RACE_ETHN_1999_2019_ESTIMATES_Y2023M07D03. We then read it into a dataframe in R. Since the data is already very clean, we simply selected the columns location_id, location_name, race_group, year_id, val, lower, upper to keep in our cleaned dataset and discarded the rest. We chose to discard those columns because every observation had the same value. For example, every observation was Female in the age group 10 to 54. We then saved this cleaned dataframe as a .Rdata file in our repository.

### Variable Descriptions

``` {r}
variable_info <- data.frame(
  Variable = c("location_id", "location_name", "race_group", "year_id", "val", "lower", "upper"),
  Data_Types = c("number", "text", "text", "number", "number", "number", "number"),
  Description = c(
    "A unique numeric identifier for the location generated and stored in an IHME database of data dimensions.",
    "The location for the estimate.",
    "The racial and ethnic group for the estimate.",
    "The time period for the estimate.",
    "Mean estimate.",
    "97.5% percentile estimate.",
    "2.5% percentile estimate."
  ),
  stringsAsFactors = FALSE
)

print(variable_info)
```

### Exploratory Analysis

```{r figure 1}
suppressPackageStartupMessages(library(tidyverse))
load("dataset/mmr.RData")

mmr_data_clean |> 
  ggplot(aes(x = val)) + 
    geom_histogram() + 
    xlim(0, 200)
```

Figure 1: The graph visualizes the frequency distribution of value MMR. From the histogram, we observe that the data is right-skewed, meaning that the majority of the data points fall on the lower end of the value range, and there are fewer observations as the value increases. This is indicated by the high bars on the left and the bars gradually decreasing in height as we move to the right, showing that high values are less common.

```{r figure 2}
mmr_data_clean |>
  group_by(year_id, race_group) |>
  summarize(mean = mean(val)) |>
  ggplot(aes(x = year_id, y = mean)) +
  geom_line(aes(color = race_group))
```

Figure 2: The line graph presented displays the trends of mean MMR across different racial and ethnic groups over a span of approximately two decades, starting from the year 2000. Each colored line on the graph represents a different racial or ethnic category, plotting the mean of 'val' for that group against the timeline of 'year_id'. Several key trends and patterns can be discerned from the graph: - The 'Non-Hispanic Black' group exhibits a notable upward trajectory, suggesting a consistent increase in the mean of 'val' over the years. Starting from a position close to other groups in the year 2000, there is a significant rise around the year 2010, and by 2020, this group has the highest mean value among all groups shown. - 'All racial and ethnic groups' combined show a gradual increase, hinting at an overall rising trend in the mean of 'val' across the population. - The 'Non-Hispanic Asian, Native Hawaiian, or Other Pacific Islander' group shows an increase, albeit with some fluctuations and a less steep incline compared to the 'Non-Hispanic Black' and 'Non-Hispanic White' groups.

```{r figure 3}
mmr_data_clean |>
  group_by(year_id, location_name) |>
  filter(str_detect(location_name, "Census")) |> 
  summarize(mean = mean(val)) |>
  ggplot(aes(x = year_id, y = mean)) +
  geom_line(aes(color = location_name))
```

Figure 3: The line graph displayed appears to represent the trends in the mean value of MMR over time across four different U.S. Census regions: Midwest, Northeast, South, and West, from the year 2000 to just before 2020. All regions show an initial increase in the mean value from 2000, with varying rates and fluctuations. Around 2005 to 2008, the Northeast region displays a sharper increase, while the Midwest shows a more gradual rise. Post-2010, all regions experience a significant increase in mean values, with the West and South regions peaking around 2015. The Northeast region exhibits a pronounced peak shortly after 2015, followed by a decline, indicating a possible reversal of the trend or a response to an event or policy change. The trends suggest regional variations in the variable measured, with each region experiencing unique patterns of change over the two decades.

```{r figure 4}
mmr_data_clean |>
  group_by(year_id, location_name, race_group) |>
  filter(str_detect(location_name, "Census")) |> 
  summarize(mean = mean(val))|>
  ggplot(aes(x = year_id, y = mean)) +
  geom_line(aes(color = location_name)) + 
  facet_wrap(~race_group)
```

Figure 4: The five graphs depict the trajectory of maternal mortality rates over a 20-year period, differentiated by racial categories and geographic regions in the United States.  These categories include Hispanic and any race, Non-Hispanic American Indian and Alaska Native, Non-Hispanic Asian and Native Hawaiian or Other Pacific Islander, Non-Hispanic Black, and Non-Hispanic White.  While there is an overarching increase in mortality rates across these racial demographics, the patterns of increase vary notably between them.

For the categories of Hispanic and any race, Non-Hispanic Asian and Native Hawaiian or Other Pacific Islander, and Non-Hispanic White, the mortality rates appear to be relatively stable, with slight fluctuations over the two decades.  The lines on the graph for these groups are relatively flat, suggesting less volatility in their yearly rates and possibly better outcomes when compared to the other categories.

Contrastingly, the Non-Hispanic Black category exhibits a pronounced upward trend in the Northeast and South regions, indicating a concerning rise in maternal mortality rates that surpasses the other racial groups in these areas.  This suggests that Non-Hispanic Black women in these regions are disproportionately affected by factors contributing to maternal mortality.

The Non-Hispanic American Indian and Alaska Native category demonstrates a significant increase in maternal mortality rates, particularly in the Midwest and West regions.  The sharp rise could be indicative of systemic health disparities or regional deficiencies in healthcare access or quality that particularly impact this demographic.

The observed regional differences, such as higher rates in the South and variations in the Midwest and West, might be influenced by a multitude of factors including healthcare infrastructure, socioeconomic status, accessibility of prenatal and postnatal care, and population density.  Regions with higher population densities might face different healthcare challenges than less densely populated areas, potentially affecting the availability and quality of care.
