---
title: "Blog Post 7"
author: "Group 13"
date: "2024-04-26"
date-modified: "2024-04-26"
draft: FALSE
---

Thesis: First of all, according to our regression graph of MMR vs.
Health Care Spending by Years.As the government spending on Health Care
increases, the MMR decreases.

As time passes, the maternal mortality rate increases across the U.S.,
and increasing spending on healthcare will positively decrease the
maternal mortality rate.

To polish our visualization graphs, we are going to add more detail and
annotation to the existing graph. We want to add a map of the US states
with fill by MMR, with a slidey bar for changing the year.

First we can add clear, descriptive titles and captions to your graphs
to convey their purpose and provide context. Annotations and highlights
are crucial for pointing out key insights or significant data points,
which can be achieved using `geom_text` or `geom_label` in ggplot2,
supported by `ggpubr`. For interactive elements, consider using the
`plotly` or `leaflet` packages for dynamic features like a sliding bar
to explore changes over time. Ensure your color choices are accessible
and consistent, and utilize `ggpubr` to adjust themes, scales, and
legends to make the graphs visually appealing and easy to understand.

```{r graph 1}
library(ggplot2)
library(readr)
library(dplyr)

data <- read_csv("dataset/Cleaned_Pregnancy_Data.csv")

average_data <- data %>%
  group_by(year, region) %>%
  summarize(MeanAge = mean(average_age_of_pregnancy, na.rm = TRUE), .groups = 'drop')

ggplot(average_data, aes(x = year, y = MeanAge, color = region, group = region)) +
  geom_line() +
  labs(title = "Trend of Mean Age of Pregnancy Over Time by Region",
       x = "year",
       y = "Mean Age of Pregnancy") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

Graph 1: This graph displays the trend of the mean age of pregnancy over
time, categorized by region. From the graph, we can figure out that: In
every region, the mean age of pregnancy has increased over time. This
suggests that, on average, women are having children at an older age
than they did in the past. The 'West' region shows the highest mean age
of pregnancy over the entire timeline, starting at around 24 years in
the 1970s and rising to just below 30 by the 2010s. The 'Midwest' and
'South' regions appear to have a similar starting point in the 1970s but
diverge slightly over time. The 'South' follows closely with the 'West',
but it still has a lower mean age of pregnancy by the 2010s. The
'Northeast' region begins with the lowest mean age of pregnancy but
increases at a rate similar to the 'Midwest', surpassing it around the
mid-2000s.

```{r graph 2}
library(ggplot2)
library(readr)

data <- read_csv("dataset/phc_mmr_with_age.csv")

data$spending <- as.numeric(data$spending)
data$AverageAge <- as.numeric(data$AverageAge)

ggplot(data, aes(x = spending, y = AverageAge, color = Region)) +
  geom_point(aes(size = 3), alpha = 0.6) +  
  geom_smooth(method = "lm", se = FALSE, aes(group = Region)) +  
  scale_size_identity() +  
  labs(title = "Bubble Chart with Fitted Lines by Region",
       x = "Healthcare Spending",
       y = "Average Age of Pregnancy",
       size = "Region") +
  theme_minimal() +
  facet_wrap(~Region)
  guides(color = guide_legend(title = "Region"))

# Print the plot
print(ggplot(data, aes(x = spending, y = AverageAge, color = Region)) +
  geom_point(aes(size = 3), alpha = 0.6) +  
  geom_smooth(method = "lm", se = FALSE, aes(group = Region)) +  
  scale_size_identity() +  
  labs(title = "Healthcare Spending vs. Average Age of Pregnancy by Region",
       x = "Healthcare Spending",
       y = "Average Age of Pregnancy") +
  theme_minimal() +
  facet_wrap(~Region)) 
  
```

Graph 2: This scatter plot compares healthcare spending with the average
age of pregnancy by region. Key observations include: There appears to
be a positive correlation between healthcare spending and the average
age of pregnancy in all regions. This implies that as healthcare
spending increases, the average age of pregnancy tends to be higher. The
data for the 'West' region, shown in purple, indicates a particularly
strong positive correlation, suggesting that increased healthcare
spending in this region might be closely associated with women having
children at older ages. Overall, the graph suggests a trend where
regions with higher healthcare spending are associated with older
average ages of pregnancy, which could reflect a variety of factors such
as access to reproductive health services, birth control, career
opportunities, or other socioeconomic factors that allow or encourage
women to delay pregnancy.

```{r graph 3}
ggplot(data, aes(x = AverageAge, y = mean_val, size = spending)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~Region) +
  labs(title = "Bubble Chart of Average Age, MMR, and Healthcare Spending",
       x = "Average Age of Pregnancy",
       y = "MMR (per 100,000 live births)",
       size = "Healthcare Spending") +
  theme_minimal()
```

Graph 3: This bubble chart illustrates the relationship between the
average age of pregnancy, maternal mortality ratio (MMR per 100,000 live
births), and healthcare spending by region. The x-axis represents the
average age of pregnancy, the y-axis represents the MMR, and the bubble
size corresponds to healthcare spending, with larger bubbles indicating
higher spending.

Observations from the chart: - There appears to be a concentration of
data points between the ages of 24 and 30, with varying MMR across the
regions. - The Midwest, Northeast, and South regions have clusters with
a wide range of MMR values but are generally centered around the same
average age of pregnancy. - The West region shows a cluster with a lower
MMR in comparison to other regions for a similar average age of
pregnancy.

The chart suggests complexity in the relationship between healthcare
spending, the average age of pregnancy, and MMR, and it would likely
require a deeper analysis to understand the underlying factors and
dynamics.


``` {r}
library(dplyr)
suppressPackageStartupMessages(library(tidyverse))
library(ggplot2)
library(dplyr)
library(ggmap)
library(patchwork)

mapp <- read_csv(here::here("dataset/phc_mmr_combined.csv"), col_types = cols(
  location_name= col_character(),
  year_id = col_double(),
  spending = col_double(),
  mean_val = col_double())) 

states_map <- map_data("state")
spending_1999 <- mapp %>% 
   filter(year_id == 9) %>% 
   mutate(state = tolower(location_name)) 
 
 map_spending <- merge(states_map, spending_1999, by.x = "region", by.y = "state", all.x = TRUE)
 
 # Create the plot
 ggplot(map_spending, aes(x = long, y = lat, group = group, fill = spending)) +
   geom_polygon(color = "white") +
   scale_fill_viridis_c(option = "C", direction = -1) +
   coord_fixed(1.3) +  
   labs(fill = "spending") +
   theme_void() 
```
 
#### Interactive Geospatial Mapping

Below you can find the code for interactive geospatial maps that allow you to view a visual of our variables of interest over time. It doesn't work in this blog post due to incompatibilities between shiny and quarto markdown, however it should work for our final pages. 
 
```{r}

library(shiny)
library(dplyr)
library(ggplot2)
library(ggmap)
library(patchwork)
library(readr)
library(here)

mapp <- read_csv(here::here("dataset/phc_mmr_combined.csv"), col_types = cols(
  location_name = col_character(),
  year_id = col_double(),
  spending = col_double(),
  mean_val = col_double()
))

states_map <- map_data("state")

ui <- fluidPage(
  titlePanel("MMR Map by Year"),
  sidebarLayout(
	sidebarPanel(
  	sliderInput("year", "Year", min = 1999, max = 2019, value = 1999, step = 1)
	),
	mainPanel(
  	plotOutput("mapPlot")
	)
  )
)

server <- function(input, output) {
  output$mapPlot <- renderPlot({
	mmr_year <- mapp %>%
  	filter(year_id == input$year) %>%
  	mutate(state = tolower(location_name))
 	 
	map_mmr <- merge(states_map, mmr_year, by.x = "region", by.y = "state", all.x = TRUE)
    

	ggplot(map_mmr, aes(x = long, y = lat, group = group, fill = mean_val)) +
  	geom_polygon(color = "white") +
  	scale_fill_viridis_c(option = "C", direction = -1) +
  	coord_fixed(1.3) +
  	labs(fill = "MMR Value") +
  	theme_void()
  })
}


shinyApp(ui = ui, server = server)

```

[Click here to visit the MMR Map by Year!](https://ma415-group13.shinyapps.io/MMR_Map_by_year/)