---
title: "Blog Post 6"
author: "Team 13"
date: "2024-04-19"
date-modified: "2024-04-19"
draft: FALSE
---
### Exploratory Data Analysis
Our exploratory data contains detailed data on personal healthcare spending on a per capita basis by state and by service from 1991 to 2020. Here are some ideas related to maternal mortality ratio based on this information:

  1. By examining the trend of per capita spending across years, one could identify which states have seen the most significant increases or decreases in spending and how the distribution of spending across services has changed.
  2. Relation to Maternal Mortality Ratio (MMR):
  
  - Access to Care: Correlating per capita spending with MMR could reveal if higher spending correlates with lower maternal mortality rates, suggesting that access to care might be a significant factor.
    
  - Funding for Maternal Health: Evaluating whether specific funding (like Medicaid expansions) is associated with improvements in maternal outcomes.
    
  - Rate Idea: Efficiency of Spending: Analyzing if increased spending in certain healthcare services (such as prenatal and postnatal care) is more effective in reducing MMR compared to others.

The rating of these ideas would depend on a detailed analysis of the CSV data against reliable maternal mortality statistics. One would need to perform statistical analysis, like regression models, to identify significant predictors of maternal mortality and to determine the strength of the relationship between healthcare spending and maternal health outcomes.

### Thesis

### Visulization

``` {r}
load("dataset/phc_mmr.RData")

combined_data |> 
  ggplot(aes(x = spending, y = mean_val)) + 
  geom_point(aes(alpha = 0.01))+  
  theme(legend.position = "none") + 
  labs(
    title = "Maternal Mortality v.s. Healthcare Spending",
    x = "Spending",
    y = "MMR"
  )
```

An initial plot of personal healthcare spending versus maternal mortality ratio shows a roughly positive correlation between the two variables which contradicts what we expected the relationship to be coming into this analysis. We would have expected a negative correlation, implying that as states spend more on healthcare overall, maternal mortality rates would be lower.

```{r}
combined_data |> 
  ggplot(aes(x = spending, y = mean_val)) + 
  geom_point(aes(alpha = 0.01)) + 
  facet_wrap(~year_id) +  
  theme(legend.position = "none") + 
  labs(
    title = "Maternal Mortality v.s. Healthcare Spending by Year",
    x = "Spending",
    y = "MMR"
  )
```

We then faceted the initial graph by year to show the trend over time. This reveals a few things. Firstly, when faceted by year the relationship no longer appears to have a positive correlation. The trend is hard to see without a regression however it does appear to be negatively correlated now. Additionally, the points for each state are spreading out. In the first few years, all the points are clumped together with lower values. However, as the years go by the points not only overall increase in value but move farther apart. This prompted us to do a standard deviation analysis to quantify this obsserved trend. 

