---
title: "Blog Post 3"
author: "Group 13"
date: "2024-03-29"
date-modified: "2024-03-29"
draft: FALSE
---
## Data Equity Analysis
In the context of maternal mortality data disaggregated by race and ethnicity, transparency is critical in understanding how racial and ethnic disaggregation is defined and applied, data sources, and data collection methods that may influence the interpretation of results. 

### Transparency
Dataset has provided detailed documentation of data sources, methods, and acknowledges any gaps or uncertainties in the data. Limitations may include incomplete data due to underreporting, differences in data collection methods between states or agencies, and challenges in capturing the complexity of racial and ethnic identity in a standardized manner. Transparency of these limits is essential to prevent misuse of data, such as oversimplified narratives that may mask the underlying causes of disparities in maternal mortality. 

### Beneficence
Good dataset require maximization of benefits and minimization of harm. For maternal mortality data, this means ensuring that data is collected, analysed and disseminated without prejudice to the communities represented, and that survey results are used equitably to improve maternal health outcomes. To uphold this principle, one limitation of practicing philanthropy may be the challenge of balancing the need for detailed data to identify differences with the risk of stigmatizing specific communities. 

### Justice
Justice involves a fair distribution of benefits and burdens. These limitations can be structural barriers to the equitable implementation of interventions, such as systemic racism in the health care system and socioeconomic inequalities that affect health outcomes.


### Initial Data Exploration

```{r figure 1}
suppressPackageStartupMessages(library(tidyverse))
load("dataset/mmr.RData")

mmr_data_clean |> 
  ggplot(aes(x = val)) + 
    geom_histogram() + 
    xlim(0, 200)
```
The graph visualizes the frequency distribution of value MMR.  From the histogram, we observe that the data is right-skewed, meaning that the majority of the data points fall on the lower end of the value range, and there are fewer observations as the value increases. This is indicated by the high bars on the left and the bars gradually decreasing in height as we move to the right, showing that high values are less common.


```{r figure 2}
mmr_data_clean |>
  group_by(year_id, race_group) |>
  summarize(mean = mean(val)) |>
  ggplot(aes(x = year_id, y = mean)) +
  geom_line(aes(color = race_group))
```
The line graph presented displays the trends of mean MMR across different racial and ethnic groups over a span of approximately two decades, starting from the year 2000. Each colored line on the graph represents a different racial or ethnic category, plotting the mean of 'val' for that group against the timeline of 'year_id'. Several key trends and patterns can be discerned from the graph: - The 'Non-Hispanic Black' group exhibits a notable upward trajectory, suggesting a consistent increase in the mean of 'val' over the years. Starting from a position close to other groups in the year 2000, there is a significant rise around the year 2010, and by 2020, this group has the highest mean value among all groups shown. - 'All racial and ethnic groups' combined show a gradual increase, hinting at an overall rising trend in the mean of 'val' across the population. - The 'Non-Hispanic Asian, Native Hawaiian, or Other Pacific Islander' group shows an increase, albeit with some fluctuations and a less steep incline compared to the 'Non-Hispanic Black' and 'Non-Hispanic White' groups.


```{r figure 3}
mmr_data_clean |>
  group_by(year_id, location_name) |>
  filter(str_detect(location_name, "Census")) |> 
  summarize(mean = mean(val)) |>
  ggplot(aes(x = year_id, y = mean)) +
  geom_line(aes(color = location_name))
```
The line graph displayed appears to represent the trends in the mean value of MMR over time across four different U.S. Census regions: Midwest, Northeast, South, and West, from the year 2000 to just before 2020. All regions show an initial increase in the mean value from 2000, with varying rates and fluctuations. Around 2005 to 2008, the Northeast region displays a sharper increase, while the Midwest shows a more gradual rise. Post-2010, all regions experience a significant increase in mean values, with the West and South regions peaking around 2015. The Northeast region exhibits a pronounced peak shortly after 2015, followed by a decline, indicating a possible reversal of the trend or a response to an event or policy change. The trends suggest regional variations in the variable measured, with each region experiencing unique patterns of change over the two decades. 
