---
title: "Blog Post 3"
author: "Christy"
date: "2024-03-29"
date-modified: "2024-03-29"
draft: FALSE
---
## Data Equity Analysis
In the context of maternal mortality data disaggregated by race and ethnicity, transparency is critical in understanding how racial and ethnic disaggregation is defined and applied, data sources, and data collection methods that may influence the interpretation of results. 

### Transparency
Dataset has provided detailed documentation of data sources, methods, and acknowledges any gaps or uncertainties in the data. Limitations may include incomplete data due to underreporting, differences in data collection methods between states or agencies, and challenges in capturing the complexity of racial and ethnic identity in a standardized manner. Transparency of these limits is essential to prevent misuse of data, such as oversimplified narratives that may mask the underlying causes of disparities in maternal mortality. 

### Beneficence
Good dataset require maximization of benefits and minimization of harm. For maternal mortality data, this means ensuring that data is collected, analysed and disseminated without prejudice to the communities represented, and that survey results are used equitably to improve maternal health outcomes. To uphold this principle, one limitation of practicing philanthropy may be the challenge of balancing the need for detailed data to identify differences with the risk of stigmatizing specific communities. 

### Justice
Justice involves a fair distribution of benefits and burdens. These limitations can be structural barriers to the equitable implementation of interventions, such as systemic racism in the health care system and socioeconomic inequalities that affect health outcomes.


### Initial Data Exploration

```{r figure 1}
suppressPackageStartupMessages(library(tidyverse))
load("dataset/mmr.RData")

mmr_data_clean |> 
  ggplot(aes(x = val)) + 
    geom_histogram() + 
    xlim(0, 200)
```


```{r figure 2}
mmr_data_clean |>
  group_by(year_id, race_group) |>
  summarize(mean = mean(val)) |>
  ggplot(aes(x = year_id, y = mean)) +
  geom_line(aes(color = race_group))
```

```{r figure 3}
mmr_data_clean |>
  filter(location_name == "Massachusetts"| location_name == "Alabama") |> 
  group_by(year_id, location_name) |>
  summarize(mean = mean(val)) |>
  ggplot(aes(x = year_id, y = mean)) +
  geom_line(aes(color = location_name))
```
